---
title: "Soc723 - TE Chapter 07"
author: "Turgut Keskintürk"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    toc: true
    number_sections: false
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.align = "center")

# load the relevant packages
pacman::p_load(
  tidyverse, purrr, causaldata, patchwork, hrbrthemes, knitr, dagitty, ggdag
)
theme_set(theme_ipsum_rc()) # theme set for the plots

```

Hello there! Let's start our exercises for Huntington-Klein's *The Effect*, Chapter 7.

## Q1

Here are the responses to the questions:

a) This is **true**. A nice causal effect on viewership is the existence of celebrities.
b) This is **false**. The diagram should include just one variable (e.g., `Evening`), with those values as responses.
c) This is **false**. The fact that we cannot measure something is not a reason for not including it.
d) This is **false**. The information is irrelevant for the DAG.

## Q2

OK, here is the DAG with the features described in the question:

```{r, fig.width = 8, fig.height = 8, dpi = 300}

dag <- dagitty(
  "
  dag
  {
  LongShift [exposure]
  LowQualityCare [outcome]
  PolicyPreferences [unobserved]
  LongShift -> LowQualityCare
  LongShift -> DoctorTired
  LongShift -> DoctorExperience
  DoctorTired -> LowQualityCare
  DoctorExperience -> LowQualityCare
  Hospital -> LongShift
  Hospital -> HospitalFunding
  Hospital -> HospitalCrowded
  HospitalFunding -> LowQualityCare
  HospitalCrowded -> LowQualityCare
  PoliciesShift -> LongShift
  PoliciesCares -> LowQualityCare
  PolicyPreferences -> PoliciesShift
  PolicyPreferences -> PoliciesCares
  }
  "
)

set.seed(11235)
ggdag_status(dag, text_col = "white", text_size = 1.4) + 
  scale_color_manual(values = c("#720500", "#244224", "#3A5874"),
                     labels = c("Treatment", "Unobserved", "Outcome")) +
  labs(color = "Nodes") +
  theme_dag(legend.position = "top")

```

## Q3

Here are the responses to the questions:

a) The **treatment** is `Funding`, while the **outcome** is `Achievement`.
b) Here are some relevant variables (wanted to keep it short, sorry but also not sorry): `GeographicRegion`, `StudentSES` and `LaggedFunding`. Let us assume that all of these variables can be observed.
c) Here are the shared variables: `GeographicRegion` and `LaggedFunding`.
d) These are *confounders*, so we want to make sure they stay in the DAG and will be adjusted for later.
e) Here is the causal diagram:

```{r, fig.width = 8, fig.height = 6, dpi = 300}

dag <- dagitty(
  "
  dag
  {
  Funding [exposure]
  Achievement [outcome]
  Funding->Achievement
  GeographicRegion->Funding
  GeographicRegion->Achievement
  StudentSES->Achievement
  GeographicRegion->StudentSES
  LaggedFunding->Funding
  LaggedFunding->Achievement
  GeographicRegion->LaggedFunding
  }
  "
)

set.seed(11235)
ggdag_status(dag, text_col = "white", text_size = 1.4) + 
  scale_color_manual(values = c("#720500", "#3A5874"),
                     labels = c("Treatment", "Outcome")) +
  labs(color = "Nodes") +
  theme_dag(legend.position = "top")

```

f) OK, it seems that the inclusion of `StudentSES` is not important to explore the causal effect, as conditioning on others is enough.


```{r, fig.width = 8, fig.height = 6, dpi = 300}

dag <- dagitty(
  "
  dag
  {
  Funding [exposure]
  Achievement [outcome]
  Funding->Achievement
  GeographicRegion->Funding
  GeographicRegion->Achievement
  LaggedFunding->Funding
  LaggedFunding->Achievement
  GeographicRegion->LaggedFunding
  }
  "
)

set.seed(11235)
ggdag_status(dag, text_col = "white", text_size = 1.4) + 
  scale_color_manual(values = c("#720500", "#3A5874"),
                     labels = c("Treatment", "Outcome")) +
  labs(color = "Nodes") +
  theme_dag(legend.position = "top")

```

## Q4

Here are the reasons why we do DAG simplifications:

a) **Unimportance**: when the causal effect in question is not important or it is negligible.
b) **Redundancy**: when the variables occupy the same space. For instance, in the example above, we can change `GeographicRegion` to two distinct variables `GeographicSES` and `GeographicSchoolN`, though the implications would remain the same.
c) **Mediators**: if the variable is just a mediator, we can drop it - though I do not agree that we should drop it. Why the hell?
d) **Irrelevance**: As our interest is on the causal path from the treatment to the outcome, we do not need to keep anything that does not affect that relationship - e.g., the `StudentSES` example above, which I dropped.

## Q5

Two ways: (1) adding a time dimension (just subscript everything with $t$) or (2) finding a source of randomness in the variable.

## Q6

Alright, these days I am thinking about the transmission of cultural beliefs from parents to teens. 

a) Let's call the cause `ParentBeliefs` and the outcome `ChildBeliefs`. This is not a "treatment" per se, but hey, it is life!
b) I am going to write 5 variables (of course) for the data generating process: `GeneticBackground` (lol, unobserved), `FamilyEnvironment`, `SchoolEnvironment`, `ParentalEducation`, and `ReligiousAffiliation`.
c) Let's draw the DAG (with some silly mistakes, so I do not have to wait for 15 minutes):

```{r, fig.width = 8, fig.height = 6, dpi = 300}

dag <- dagitty(
  "
  dag
  {
  ParentBeliefs [exposure]
  ChildBeliefs [outcome]
  GeneticBackground [unobserved]
  ParentBeliefs -> ChildBeliefs
  ParentBeliefs <- GeneticBackground -> ChildBeliefs
  ParentBeliefs <- FamilyEnvironment -> ChildBeliefs
  ParentBeliefs <- ReligiousAffiliation -> ChildBeliefs
  ParentBeliefs <- ParentalEducation -> ChildBeliefs
  SchoolEnvironment -> ChildBeliefs
  }
  "
)

set.seed(11235)
ggdag_status(dag, text_col = "white", text_size = 1.25) + 
  scale_color_manual(values = c("#720500", "#244224", "#3A5874"),
                     labels = c("Treatment", "Unobserved", "Outcome")) +
  labs(color = "Nodes") +
  theme_dag(legend.position = "top")

```

d) OK, I waited for 15 minutes. The bad assumption: `ParentalEducation` should affect `SchoolEnvironment` and `ReligiousAffiliation`, as it is highly probably that parents with differing levels of education are differentiated along these variables.

## Q7

I am going to add the diagram the time dimension, so that it makes some sense:

```{r, fig.width = 15, fig.height = 5, dpi = 300}

dag <- dagitty(
  "
  dag
  {
  MotivationT1 -> MotivationT2
  AchievementT1 -> AchievementT2
  MotivationT1 -> AchievementT2
  AchievementT1 -> MotivationT2
  }
  "
)

set.seed(11235)
ggdag_classic(dag, layout = "grid") + theme_dag()

```

Thank you!
