---
title: "Soc722 - SR Chapter 3"
author: "Turgut Keskintürk"
date: '2022-09-23'
output:
  html_document:
    theme: united
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hello there! Let's start with our exercises for McElreath's *Statistical Rethinking*, Chapter 3.

# Set-Up

```{r, message=FALSE}

# load the relevant packages
library(tidyverse) # again, I silenced this chunk to exclude messages.
library(rethinking)

```

# Responses to Questions

## Easy Problems

I use McElreath's code block (3.27) to have a specific set of samples first, per instructions. I change certain parts to tidyverse format. I am not sure if ``dplyr`` sampling is exactly the same to base ``sample``, so for this particular example, I used the base version.

```{r}

# construct the data

d <- tibble(p_grid = seq(
  from = 0,
  to = 1,
  length.out = 1000
),
prior = rep(1, 1000)) %>%
  mutate(likelihood = dbinom(6, size = 9, prob = p_grid)) %>%
  mutate(posterior = likelihood * prior / sum(likelihood * prior))

# set seed

set.seed(100)

# draw some good old samples

samples <- tibble(samples = sample(
  d$p_grid,
  prob = d$posterior,
  size = 1e4,
  replace = TRUE
))

```

**3E1**. Posterior probability that lies below p = 0.2.

```{r}

samples %>%
  filter(samples < 0.2) %>%
  summarise(sum = n() / 1e4)

```

The answer is **0.0004**. *Very low!*

**3E2**. Posterior probability that lies above p = 0.8.

```{r}

samples %>%
  filter(samples > 0.8) %>%
  summarise(sum = n() / 1e4)

```

The answer is **0.112**

**3E3**. Posterior probability that lies between p = 0.2 and p = 0.8.

```{r}

samples %>%
  filter(samples > 0.2 & samples < 0.8) %>%
  summarise(sum = n() / 1e4)

```

Well, we would expect to see something that nearly (the exact points excluded though) sums to 1, so the answer is **0.888.** OK, we now turn to the questions that ask us to compute the opposite of what we had above.

**3E4**. 20% of the posterior probability lies below:

```{r}

samples %>% 
  summarise(`20th percentile` = quantile(samples, p = .2))

```

We see that the answer is **0.519**.

**3E5**. 20% of the posterior probability lies above:

```{r}

samples %>% 
  summarise(`80th percentile` = quantile(samples, p = .8))

```

We see that the answer is **0.756**.

**3E6**. The question asks of values of *p* that contain the narrowest interval equal to 66% of posterior probability. 

```{r}

rethinking::HPDI(samples$samples, prob = 0.66)

```

**3E7**. The question asks of values of *p* that contain 66% of the posterior probability, assuming equality below and above the interval.

```{r}

rethinking::PI(samples$samples, prob = 0.66)

```

Nice tricks! We have completed the easy problems.

## Medium Problems

Let's clean up the environment before proceeding to the next set of questions (my motto: best environment is a clean environment).

```{r}

rm(list = ls())

```

The next set of questions wants us to assume that globe tossing data end up with 8 water in 15 tosses.

**3M1**. Let's construct the posterior distribution using grid approximation.

```{r}

d <- tibble(p_grid = seq(
  from = 0,
  to = 1,
  length.out = 1000
),
prior = rep(1, 1000)) %>%
  mutate(likelihood = dbinom(8, size = 15, prob = p_grid)) %>%
  mutate(posterior = likelihood * prior / sum(likelihood * prior))

```

**3M2**. Nice, we have the posterior distribution with flat priors now. Let's now draw 10,000 (=1e4) samples from this approximation. Then, we are asked to return the 90% HPDI for *p*.

```{r}

# sampling

set.seed(112358)

samples <-
  d %>%
  dplyr::slice_sample(n = 1e4, weight_by = posterior, replace = T)

# hpdi

rethinking::HPDI(samples$p_grid, prob = 0.9)

```

Nicely done!

**3M3**. We are going to construct a posterior predictive check for the data. In doing this, I will (a) simulate the distribution, (b) averaging over the posterior uncertainty in *p*. We are asked for Pr(8 Water in 15 Tosses).

```{r}

# create a dummy data with 100,000 observations

set.seed(112358)

dummy <- tibble(draws = rbinom(1e5, size = 15, prob = samples$p_grid))

# plot

dummy %>%
  ggplot(aes(x = draws)) + 
  geom_histogram(binwidth = 0.5) +
  theme_bw() +
  labs(x = "Pr(W)", y = "Frequency")

# probability of observing 8 tosses

dummy %>%
  filter(draws == 8) %>%
  summarize(Prob = n() / 1e5)

```

We see that the probability of observing 8 tosses out of 15 is **0.148**.

**3M4**. Now, we are gonna look at the Pr(6 Water in 9 Tosses).

```{r}

# create a dummy data with 100,000 observations

set.seed(112358)

dummy <- tibble(draws = rbinom(1e5, size = 9, prob = samples$p_grid))

# plot

dummy %>%
  ggplot(aes(x = draws)) + 
  geom_histogram(binwidth = 0.25) +
  theme_bw() +
  labs(x = "Pr(W)", y = "Frequency")

# probability of observing 6 tosses

dummy %>%
  filter(draws == 6) %>%
  summarize(Prob = n() / 1e5)

```

Cool! The answer is **0.174**.

**3M5**. OK, we are now changing our priors. Exciting! If *p* < 0.5, we are gonna use 0, and constant otherwise. Everything is same.

```{r}

# clean-up again

rm(list = ls())

# new distribution

d <- tibble(
  p_grid = seq(
    from = 0,
    to = 1,
    length.out = 1000
  ),
  prior = ifelse(p_grid <= 0.5, 0, 1)
) %>%
  mutate(likelihood = dbinom(8, size = 15, prob = p_grid)) %>%
  mutate(posterior = likelihood * prior / sum(likelihood * prior))

# sampling

set.seed(112358)

samples <-
  d %>%
  dplyr::slice_sample(n = 1e4,
                      weight_by = posterior,
                      replace = T)
```

Nice. We updated everything. Let's look at the HPDI, and the remaining plots.


```{r}

# hpdi

rethinking::HPDI(samples$p_grid, prob = 0.9)

# 8 out of 15 tosses

## create a dummy data with 100,000 observations

set.seed(112358)

dummy <- tibble(draws = rbinom(1e5, size = 15, prob = samples$p_grid))

## plot

dummy %>%
  ggplot(aes(x = draws)) + 
  geom_histogram(binwidth = 0.5) +
  theme_bw() +
  labs(x = "Pr(W)", y = "Frequency")

## probability of observing 8 tosses

dummy %>%
  filter(draws == 8) %>%
  summarize(Prob = n() / 1e5)

# 6 out of 9 tosses

## create a dummy data with 100,000 observations

set.seed(112358)

dummy <- tibble(draws = rbinom(1e5, size = 9, prob = samples$p_grid))

## plot

dummy %>%
  ggplot(aes(x = draws)) + 
  geom_histogram(binwidth = 0.25) +
  theme_bw() +
  labs(x = "Pr(W)", y = "Frequency")

## probability of observing 6 tosses

dummy %>%
  filter(draws == 6) %>%
  summarize(Prob = n() / 1e5)

```

OK, we see that HPDI is pushed to the right. Similar observations can be made with regard to the probabilities (8 out of 15 is pushed to 0.148 to 0.159, and 6 out of 9 is pushed from 0.174 to 0.231).

**3M6**. It seems that I am gonna do some for loops. But not sure.

```{r}

# automate?

how_many_tosses <- matrix(nrow = 250, ncol = 2)
  loop_friend <- 1

for (i in seq(from = 1, to = 5000, by = 20)) {
  precise_rbinoms <- rbinom(1, size = i, prob = 0.7)
  precise_rbinoms <-
    tibble(p_grid = seq(from = 0, to = 1, length.out = 1000),
           prior = rep(1, 1000)) %>%
    mutate(likelihood = dbinom(precise_rbinoms, size = i, prob = p_grid)) %>%
    mutate(posterior = likelihood * prior / sum(likelihood * prior))
  precise_samples <- precise_rbinoms %>%
    slice_sample(n = 10000, weight_by = posterior, replace = TRUE)
  intervals <- rethinking::PI(precise_samples$p_grid, prob = .99)
  how_many_tosses[[loop_friend, 1]] <- i
  how_many_tosses[[loop_friend, 2]] <- intervals[[2]] - intervals[[1]]
  loop_friend <- loop_friend + 1
}

# making it easy
  
how_many_tosses <- as.data.frame(how_many_tosses)

```

Aaah, no. This makes us approximate, but for a precise answer, no. I give up.

## Hard Problems

It is now time for the hard problems. Scary. But fun maybe? Let's see.

We are going to load the data first.

```{r}

rm(list = ls())
data(homeworkch3)

# maybe a tibble data frame would work later on?

df <- tibble(
  birth1 = birth1,
  birth2 = birth2,
  b = birth1 + birth2,
  g = 2 - b
)

```

**3H1**. We are going to use grid approximation to compute the posterior distribution of a birth being boy.

```{r}

# construct the posterior distribution

d <- tibble(p_grid = seq(
  from = 0,
  to = 1,
  length.out = 100
),
prior = rep(1, 100)) %>%
  mutate(likelihood = dbinom(sum(df$b), size = 200, prob = p_grid)) %>%
  mutate(posterior = likelihood * prior / sum(likelihood * prior))

# which parameter value maximizes the posterior probability?

d %>%
  arrange(desc(posterior)) %>%
  slice(1) %>%
  print()

```

It seems that the parameter value **0.556** maximizes the posterior probability.

**3H2**. We are now going to sample 10,000 random parameters from the posterior distributions above.

```{r}

# sampling

set.seed(112358)

samples <-
  d %>%
  dplyr::slice_sample(n = 1e4, weight_by = posterior, replace = T)

# highest posterior density intervals

rethinking::HPDI(samples$p_grid, prob = 0.50)
rethinking::HPDI(samples$p_grid, prob = 0.89)
rethinking::HPDI(samples$p_grid, prob = 0.97)

```

Cool!

**3H3**. Let's go with simulation. King McElreath proposes the ``dens`` command, so no tidyverse this time, Nico!

```{r}

# draws

set.seed(112358)
dummy <- tibble(draws = rbinom(1e4, size = 200, prob = samples$p_grid))

# plot

rethinking::dens(dummy$draws)

```

Yes, it seems that the data shows that the actual observation is a central outcome.

**3H4**. Now, we are going to simulate only 100 observations, and compare this with 10,000 dummy observations.

```{r}

# recalibration

d <- tibble(p_grid = seq(
  from = 0,
  to = 1,
  length.out = 100
),
prior = rep(1, 100)) %>%
  mutate(likelihood = dbinom(sum(df$birth1), size = 100, prob = p_grid)) %>%
  mutate(posterior = likelihood * prior / sum(likelihood * prior))

# new sampling for the first borns

samples <-
  d %>%
  dplyr::slice_sample(n = 1e4, weight_by = posterior, replace = T)

# new draws

set.seed(112358)
dummy_new <- tibble(draws = rbinom(1e4, size = 100, prob = samples$p_grid))

# plot

rethinking::dens(dummy_new$draws)

```

The actual number of boys in first-borns is ``sum(birth1)`` = **51**, which is pretty close, once more, to the modal value in this plot.

**3H5**. OK, this is a bit complicated. I think we first need to know the number of girls among those who are first-born. This is easy. We can basically look it up as ``sum(birth1==0)``, which equals to 49. Now, we are going to simulate 10,000 births.

```{r}

girls_simulation <- tibble(draws = rbinom(1e4, size = 49, prob = samples$p_grid))

# same plot for McElreath

rethinking::dens(girls_simulation$draws)

```

OK, now, as you can see, the modal response turns to 25. What about the number of boys following the first-born girls. Let's learn that.

```{r}

df %>%
  filter(birth1 == 0) %>%
  summarize(sumboys = sum(birth2))

```

Huh, we see 39 boys here, compared to 25ish numbers, which suggests that the processes are not independent; namely, the first birth and the second birth has some sort of a dependence relationship in terms of biological sex!

Thank you!
