---
title: "Soc722 - DV Chapters 1-3"
author: "Turgut Keskintürk"
date: '2022-09-13'
output:
  html_document:
    theme: united
    toc: true
    number_sections: true
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hello there! Let's start with our exercises for Kieran Healy's Data Visualization, Chapter 1 to 3.

# Set-Up

```{r}

# load the best package in the world
library(tidyverse)

# load the data
exercise_data <- read_csv("./Data/visualize_data.csv")
glimpse(exercise_data)

```

# Responses to Questions

## Question 1

I believe we will see a somewhat curvilinear relationship: people who record more exercise will have less BMI, but up to a point (simply because low BMI is also a sign of problem in terms of health).  

```{r}

cor(exercise_data$Exercise, exercise_data$BMI)

```

Huh! No relation whatsoever. We have a slight negative correlation, but it is nearly non-existent.

Let's look at the data visually.

```{r}

exercise_data %>%
  ggplot(aes(Exercise, BMI)) +
  geom_point() +
  theme_bw() +
  labs(title = "The Relationship Between Exercise and BMI", 
       x = "Exercise (in Minutes)", y = "BMI")

```

Hahaha! Nice. This is awesome.

## Question 2

OK, we are now going to study with a dataset from ``causact`` package. Let's install it.

```{r, eval = FALSE}

install.packages("causact")

```

A couple of other set-ups.

```{r}

library(causact)
data(corruptDF)
glimpse(corruptDF)

```

So, ``CPI2017`` basically measures the perceived corruption in a country's public sector, while ``HDI2017`` mainly measures the degree to which a country is "developed," an average index that uses longevity, income and education.

## Question 3

Let's make a scatterplot.

```{r}

corruptDF %>%
  ggplot(aes(HDI2017, CPI2017)) +
  geom_point() +
  theme_bw() +
  labs(title = "The Relationship Between Human Development and Perceived Corruption", 
       x = "Corruption Perceptions Index", y = "Human Development Index")

```

Nice. We see that there is a positive relationship between these variables, where low corruption is correlated with high development.

## Question 4

OK, let's add a new layer.

```{r}

# method 1

corruptDF %>%
  ggplot(aes(HDI2017, CPI2017)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  labs(title = "The Relationship Between Human Development and Perceived Corruption", 
       x = "Corruption Perceptions Index", y = "Human Development Index")

# method 2

corruptDF %>%
  ggplot(aes(HDI2017, CPI2017)) +
  geom_point() +
  geom_smooth(method = "gam") +
  theme_bw() +
  labs(title = "The Relationship Between Human Development and Perceived Corruption", 
       x = "Corruption Perceptions Index", y = "Human Development Index")

```

The second method definitely fits the data better, as opposed to the linear method, which forces a linear structure to the relationship. Instead, the second method show that there is nearly a threshold value of corruption in terms of its relation to development.

## Question 5

OK, we are going to add another layer.

```{r}

corruptDF %>%
  ggplot(aes(HDI2017, CPI2017, fill = region, color = region)) +
  geom_point() +
  geom_smooth(method = "gam") +
  theme_bw() +
  labs(title = "The Relationship Between Human Development and Perceived Corruption", 
       x = "Corruption Perceptions Index", y = "Human Development Index") +
  theme(legend.title = element_blank())

```

Ugh! No way. Let's facet this graph to have a decent look.

```{r}

corruptDF %>%
  ggplot(aes(HDI2017, CPI2017)) +
  geom_point() +
  geom_smooth(method = "gam") +
  theme_bw() +
  labs(title = "The Relationship Between Human Development and Perceived Corruption", 
       x = "Corruption Perceptions Index", y = "Human Development Index") +
  facet_wrap(~ region, nrow = 2)

```

## Question 6

We are now going to reverse the scale of the X axis.

```{r}

corruptDF %>%
  ggplot(aes(HDI2017, CPI2017)) +
  geom_point() +
  geom_smooth(method = "gam") +
  theme_bw() +
  labs(title = "The Relationship Between Human Development and Perceived Corruption", 
       x = "Corruption Perceptions Index", y = "Human Development Index") +
  facet_wrap(~ region, nrow = 2) +
  scale_x_reverse()

```

Well, to be honest, this is a lot more intuitive simply because of the name of the measure!

## Question 7

Let's add a subtitle and caption.

```{r}

plot_corruptDF <- corruptDF %>%
  ggplot(aes(HDI2017, CPI2017)) +
  geom_point() +
  geom_smooth(method = "gam") +
  theme_bw() +
  labs(title = "The Relationship Between Human Development and Perceived Corruption",
       subtitle = "Observations As Countries. N = 174.",
       x = "Corruption Perceptions Index", y = "Human Development Index",
       caption = "Corruption Index comes from Transparancy International, while Development Index comes from UNDP.") +
  facet_wrap(~ region, nrow = 2) +
  scale_x_reverse()

plot_corruptDF

```

## Question 8

This is nice! Let's save the plot to as PNG file.

```{r}

ggsave("cpi_vs_hdi_plot.png", plot = plot_corruptDF, h = 7.5, w = 10, units = "in")

```

Thank you!
