---
title: "Soc723 - MD Chapter 7"
author: "Turgut Keskintürk"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.align = "center")

# load the relevant packages
pacman::p_load(tidyverse, janitor, broom, patchwork, hrbrthemes, knitr)
theme_set(theme_ipsum_rc()) # theme set for the plots

data(bikes, package = "bayesrules")

```

Hello there! Let's start with our exercises for Chapter 7.

# Question 0

Answers to the questions:

- The number of rows is **500**,
- Each row represents **a specific date of the year**,
- The data cover **days in 2011 and 2012**,
- The maximum ridership is **6,946**,
- The highest wind speed was **34**.

# Question 1

Let's start with the correlations:

```{r}

cor(bikes$rides, bikes$temp_feel) |> round(2)
cor(bikes$rides, bikes$windspeed) |> round(2)

```

As we can see, the number of rides is positively associated with the felt temperature, while negatively associated with windspeed.

# Question 2

Thank you very much for turning the miles into kilometers, as any sensible person should do!

```{r}

# mutate
bikes <- bikes |> 
  mutate(wind_kphs = 1.61 * windspeed)

# correlate
cor(bikes$windspeed, bikes$wind_kphs)

```

The correlation coefficient measures the linear association between two variables. As we can see above, the relationship between KPH and MPH is just a perfectly linear relationship (both is the linear multiplication of the other), which is why the coefficient is 1.

# Question 3

Let's do some regressions:

```{r}

fit1 <- lm(rides ~ windspeed, data = bikes)
fit2 <- lm(rides ~ wind_kphs, data = bikes)

fit1 |> tidy() |> 
  subset(select = -p.value) |> 
  mutate(term = ifelse(term == "windspeed", "Wind per Mile/Hour", term)) |> 
  kable(digits = 2, align = "lcccc")
fit2 |> tidy() |> 
  subset(select = -p.value) |> 
  mutate(term = ifelse(term == "wind_kphs", "Wind per Km/Hour", term)) |> 
  kable(digits = 2, align = "lcccc")

```

OK, the coefficients are different, as the units over which the linear model generates coefficients are different. If we would create predictions from these two models, we would end up with the same predictions over different values of windspeed.

Notice the t-stats, which are identical. This is not surprising, we just basically rescaled our variables.

# Question 4

If the wind is 20 KPH, the ridership would be **3515**. If the wind is 20 MPH, the ridership would be **3095**.

# Question 5

Alright, some data manipulations and model fits below:

```{r}

# celcius (for sensible people)
bikes <- bikes |> 
  mutate(temp_c = (temp_feel - 32)*5/9)

# model
fit <- lm(rides ~ wind_kphs + temp_c, data = bikes)

# show me
fit |> tidy() |> 
  subset(select = -p.value) |> 
  mutate(term = ifelse(term == "wind_kphs", "Wind per Km/Hour", term),
         term = ifelse(term == "temp_c", "Temperature in Celcius", term)) |> 
  kable(digits = 2, align = "lcccc")

```

- For a 1 unit increase in `wind_kphs`, the model expects a 19.8 decrease in `rides`.
- For a 1 unit increase in `temp_c`, the model expects a 143 increase in `rides`.

The intercept defines that weird universe where the wind per kilometer/hour is 0 and the temperature is also 0. Like hell.

# Question 6

Let's predict what Nico wanted:

```{r}

# prediction dataframe
pred_df <- tibble(
  situation = 1:3, temp_c = c(25, 15, 10), wind_kphs = c(15, 5, 40)
)

# predict
pred_df <- pred_df |> 
  mutate(predictions = predict(fit, newdata = pred_df))

# predictions
pred_df$predictions

```

# Question 7

Gonna add a new predictor, `weekend`, to complicate our model:

```{r}

# model
fit <- lm(rides ~ wind_kphs + temp_c + weekend, data = bikes)

# show me
fit |> tidy() |> 
  subset(select = -p.value) |> 
  mutate(term = ifelse(term == "wind_kphs", "Wind per Km/Hour", term),
         term = ifelse(term == "temp_c", "Temperature in Celcius", term),
         term = ifelse(term == "weekendTRUE", "Weekend", term)) |> 
  kable(digits = 2, align = "lcccc")

```

- If it is a weekend, the model expects a 713 decrease in `rides`.

# Question 8

Another prediction. Let's build up the dataframe and predict:

```{r}

# prediction dataframe
pred_df <- tibble(
  situation = 1:2, 
  temp_c = c(rep(mean(bikes$temp_c))), 
  wind_kphs = c(rep(mean(bikes$wind_kph))),
  weekend = c(FALSE, TRUE)
)

# predict
pred_df <- pred_df |> 
  mutate(predictions = predict(fit, newdata = pred_df))

# predictions
pred_df$predictions

```

# Question 9

We are asked to find the day with the largest residual where the model is at its worst. Every model gets that, I hear.

```{r}

# generate residuals
bikes <- bikes |> 
  mutate(finalpreds = predict(fit, newdata = bikes)) |> 
  mutate(resids = abs(finalpreds - rides))

# show me the day
bikes |> arrange(desc(resids)) |> slice(1) |> select(date)

```

There were basically **no** (20) rides in that date. I looked at the date and saw that there was Hurricane Sandy. I guess that's the reason.

Thank you!
