---
title: "Soc722 - MD Chapter 3"
author: "Turgut Keskintürk"
date: '2022-09-07'
output:
  html_document:
    theme: united
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.align = "center")
```

Hello there! Let's start with our exercises for Chapter 3.

# Set-Up

First, we will begin with loading the tidyverse package and the dataset we will use for this exercise.

```{r}

# loading the tidyverse package
library(tidyverse)

# loading the data
mario_kart <- read_csv(
  "https://raw.githubusercontent.com/vaiseys/223_course/main/Data/world_records.csv")

# have a quick look at the data
glimpse(mario_kart)

```

# Responses to Questions

Nice, we have the data. Let's start with the first question.

## Question 1

Let's first keep the records with type *Three Lap*. 

```{r}

three_laps <- mario_kart %>%
  filter(type == "Three Lap")

```

OK. We basically want to create two datasets: one without Rainbow Road track, and one that contains only that track. Let's go.

```{r}

# all tracks w/o rainbows
kart_neva_rainbows <- three_laps %>%
  filter(track != "Rainbow Road")

# only rainbow road track
kart_only_rainbows <- three_laps %>%
  filter(track == "Rainbow Road")

```

# Question 2

We are now going to compare the Rainbow Road track to the other tracks through the average time spent in the game.

```{r}

# summarize the rainbow road
kart_only_rainbows %>%
  summarize(
    mean = mean(time),
    sd = sd(time)
  )

# summarize all the others
kart_neva_rainbows %>%
  summarize(
    mean = mean(time),
    sd = sd(time)
  )


```

We see that those that played in the Rainbow Road track had significantly higher time played compared to *all the other* tracks, meaning that what the exercises said about Rainbow Road not being a cool regular track seems to hold, no?

Yet, what if there are even **worse** tracks in that big lump? Let's go find out about that.

# Question 3

We are going to do two things: (a) we are gonna look at the established records for each track & (b) find out the most achieved track.

```{r}

three_laps %>%
  group_by(track) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

```

We see that **Toad's Turnpike** has a lot of things going on!

# Question 4

OK, let's look at the drivers now.

```{r}

three_laps %>%
  group_by(player, track) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  head() # just to show the first couple of results.

```

Nice! It seems that Penev scored 26 records in the Choco Mountain track. Good job, Penev!

# Question 5

This track and that track - we don't even know the averages yet! Let's look at them.

```{r}

three_laps %>%
  group_by(track) %>%
  summarise(mean = mean(time)) %>%
  arrange(desc(mean)) %>%
  head() # just to show the first couple of results.

```

Well, yeah, Rainbow Road shines once more. Good. Let's look at the records one by one. 

```{r}

three_laps %>% 
  group_by(track) %>% 
  arrange(time) %>%
  slice(1) %>%
  select(track, time)

```

Neat!

# Question 6

We are now going to create a binary variable that takes 1 if record_duration is higher than 100 and 0 otherwise.

```{r}

three_laps <- three_laps %>%
  mutate(
    records100days = as.integer(record_duration > 100)
  )

```

OK, which one of these awesome players hold the most number of records over 100 days?

```{r}

three_laps %>%
  group_by(player) %>%
  summarize(records = sum(records100days)) %>%
  arrange(desc(records)) %>%
  head()

```

Mr. MR seems to hold the best resume.

# Question 7

Let's load the drivers dataset.

```{r}

drivers <- read_csv(
  "https://raw.githubusercontent.com/vaiseys/223_course/main/Data/drivers.csv")

```

Let's join the drivers dataset with our usual guys, using the player name. However, there are multiple observations in the drivers dataset, which naturally increases the row count with a left_join. I am going to trim the dataset first and then join.

```{r}

# trim the drivers dataset
drivers <- drivers %>%
  select(player, nation) %>%
  distinct()

# now, we are going to perform a left-join
three_laps <- three_laps %>%
  left_join(drivers, by = "player")

```

Let's plot the barchart for how many records there are.

```{r}

three_laps %>%
  drop_na() %>% # some missing observations
  group_by(nation) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = reorder(nation, n), n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_bw() +
  labs(title = "The Number of Records for Each Countries",
       x = "Nation", y = "The # of Records") +
  coord_flip()


```

Thank you!
