---
title: "Soc723 - MD Chapter 6"
author: "Turgut Keskintürk"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.align = "center")

# load the relevant packages
pacman::p_load(tidyverse, janitor, broom, patchwork, hrbrthemes, knitr)
theme_set(theme_ipsum_rc()) # theme set for the plots

```

Hello there! Let's start with our exercises for Chapter 6.

```{r}

d <- read.csv(
  "https://raw.githubusercontent.com/vaiseys/223_course/main/Data/gender_employment.csv"
) |> clean_names()

```

# Question 1

Let's fit our model:

```{r}

# some data tweaks
d <- d |> 
  mutate(
    major_category = as.factor(major_category),
    major_category = relevel(major_category, ref = "Management, Business, and Financial")
  ) |> 
  mutate(year = year - 2013) # gonna center the year at 0 to avoid bad intercept

# fit the data, and name it something other than nico's ridiculously long name
fit <- lm(wage_percent_of_male ~ year + major_category, data = d)

# show me
fit |> tidy() |> 
  select(-p.value) |> 
  mutate(term = str_remove(term, pattern = "major_category"),
         term = ifelse(term == "year", "Year", term)) |> # prettify
  kable(digits = 2, align = "lcccc")

```

OK, it seems that there is not much trend overall in terms of year, even though there is a slight (very slight!) positive take. It is insane that the Management degrees produce such an asshole wage distribution (though not insane perhaps, but predictable).

Some algebraic operations gave us the following responses: (a) **79.46** for Sales and (b) **82.4** for Service.

# Question 2

Let's see if the parallel trends assumption is warranted:

```{r, fig.width = 8, fig.height = 8, dpi = 300}

d |>
  ggplot(aes(x = year, y = wage_percent_of_male)) +
  geom_jitter(alpha = 0.1, col = "#720500") +
  geom_smooth(method = "lm", col = "black") +
  labs(title = "Women's Earnings with Respect to Men's",
       y = "% of Men's Income",
       x = "Year") +
  facet_wrap(~major_category, nrow = 4)

```

Seems like it, with the exception of `Natural Resources, Constuction, and Maintenance`, which shows an uptick for this category.

# Question 3

Let's do the interaction and inspect the results.

```{r}

# fit
int <- lm(wage_percent_of_male ~ year * major_category, data = d)

# show me
int |> tidy() |> 
  select(-p.value) |> 
  mutate(
    term = str_replace(term, pattern = "year:major_category", "Interaction: Year x "),
    term = str_remove(term, pattern = "major_category"),
    term = ifelse(term == "year", "Year", term)) |> # prettify
  kable(digits = 2, align = "lcccc")

```

Some algebraic operations gave us the following responses: (a) **96.06** for Computing and (b) **81.74** for Service. Wow!

# Question 4

We do not want to overfit our model to the data, and well, *less is better.*

# Question 5

We are still interested in how `wage_percent_of_male` changed across these years. Let's fit some models and describe:

```{r}

# fits
fit1 <- lm(wage_percent_of_male ~ year, data = d)

# show me
fit1 |> tidy() |> 
  select(-p.value) |> # this is not necessary I know, but it's fun
  mutate(term = ifelse(term == "year", "Year", term)) |> # same
  kable(digits = 2, align = "lcccc")

```

As we talked above, there is a slight uptick, but it's not much clear what's going on with this model. We are going to add `percent_female` to our model toolkit. Let's see the correlation among these variables.

```{r}

d |> 
  select(year, wage_percent_of_male, percent_female) %>% 
  cor(use = "complete.obs") |> 
  kable(digits = 2, align = "ccc")

```

OK, the correlations are pretty low, though `percent_female` might pick up some of the variation here. Let's build:

```{r}

# fits
fit2 <- lm(wage_percent_of_male ~ year + percent_female, data = d)

# show me
fit2 |> tidy() |> 
  select(-p.value) |> # this is not necessary I know, but it's fun
  mutate(term = ifelse(term == "year", "Year", term),
         term = ifelse(term == "percent_female", "% Female", term)) |> # same
  kable(digits = 2, align = "lcccc")

```

No, practically no change. This model basically means that conditioning on percent_female does not change `year` effects.

# Question 6

R-Squared is, as the term implies, the square of R (lol), which is the explained variance in the dependent variable that can be attributed to the set of independent variables used in our model. This is a sloppy definition, but it should work.

Let's compare the model $R^2$:

```{r}

glance(fit1)$r.squared
glance(fit2)$r.squared

```

Not surprisingly, the $R^2$ of the second model is higher than the first model. I do not want to conclude anything from this, to be honest. Just that `percent_women` has some sort of relationship to the outcome, and it is nice to pick that up.

Thank you!